Parallel global optimization on GPU \thanks{The revised version of the paper was supported by the Russian Science Foundation, project No 15-11-30022 ``Global optimization, supercomputing computations, and applications''}

### 2019-10-07 2016/JoGO_GPU/BarkalovGergel.tex
JOGO2016 paper next version

\title{Parallel global optimization on GPU 
	\thanks{The research is supported by the grant of the Ministry of education and science of the Russian Federation (the agreement of August 27, 2013 No 02.B.49.21.0003)}
}
%\subtitle{Do you have a subtitle?\\ If so, write it here}
%\titlerunning{Short form of title}        % if too long for running head

@@

\institute{K. Barkalov  \at
              Computing Mathematics and Cybernetics Faculty, Lobachevsky State University of Nizhni Novgorod, Nizhni Novgorod, Russia \\
              Tel.: +7-831-4623356\\
              Fax: +7-831-4623085\\
              \email{barkalov@vmk.unn.ru} 
           \and
           V. Gergel \at
							\email{gergel@unn.ru} 
}

@@

\begin{abstract}
This work considers a parallel algorithm for solving multidimensional multiextremal optimization problems. This algorithm uses Peano-type space filling curves for dimension reduction. Conditions of non-redundant parallelization of the algorithm are considered. Implementation of the algorithm on modern computing systems with the use of a graphics processing unit (GPU) is investigated. Speedup of the algorithm using GPU as compared with the same algorithm implemented on CPU only is demonstrated experimentally. Computing experiments are carried out on a series of several hundred multidimensional multiextremal problems.
\keywords{Global optimization \and Parallel computing \and Graphics processing unit \and Peano-type space filling curves \and Characteristical algorithms}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

This paper considers parallel algorithms for solving multiextremal optimization problems. The objective function in such problems is characterized by existence of several local optima (usually their number is not known and can be large). 
%new text begin
In multiextremal problems the possibility of a reliable estimation of the global optimum is essentially based on the availability of an a priori information about the function, allowing us to estimate lower bound of the objective function using known values at search points. 
%new text end
Very often, such information about the problem is presented as the assumption that the objective function $\varphi$ satisfies a Lipschitz condition with an a priori unknown constant $L$. 
%new text begin
The important feature of the problem under consideration is that the objective function is commonly not given by an analytical formula. The usual way to calculate the function value is to apply a computer-aided algorithm. Thus, each calculation may be interpreted as running some ``black-box'' to produce the outcome for the preset input. These runs could require substantial computer resources and therefore should not be too numerous. 
%new text end
%new word - such
Such problems are common in applications (problems of optimal design of objects and technological processes in various fields of technology, problems of model fitting according to observed data in scientific research, optimization of control processes, etc., see the review \cite{RefPinter}).

Multiextremal optimization problems offer little in terms of analytical solution, while numerical solution is computationally challenging as computational costs often grow exponentially in dimension. By employing modern parallel computational systems it is possible to broaden the area of applicability of global optimization methods. At the same time, the problem of efficient parallelization of the global search process arises. Therefore, development of efficient parallel methods for numerical solution of multiextremal optimization problems and their software implementation for modern computational systems is the obvious focus of research in this area.

One of the promising trends in the field of parallel global optimization (as in many areas connected with software implementation of complex algorithms) is the use of graphics processing units (GPUs). During the last decade, productivity of GPUs has been rapidly growing to satisfy the continuously growing demand of graphic application developers. Furthermore, some principles of development of graphics hardware have changed over the past few years. Today a GPU is a high-performance flexibly programmable and massively parallel processor that can provide solution to many computationally complex problems \cite{RefHwu}.

%new text begin
By now, parallel implementations have been suggested for most of the existing sequential algorithms. A review of modern methods of parallel global optimization can be found, for example, in \cite{RefApuzzo}. The methods described in \cite{RefApuzzo} can be divided into two classes: deterministic methods of global optimization, and heuristic methods, which are one way or another based on the random search concept. Algorithms belonging to the latter class allow the use of massive parallelism and are easily implemented on GPU (see works \cite{RefFerreiro,RefZhu,RefGarcia,RefMussi} and the review \cite{RefLangdon}). Owing to their stochastic nature, algorithms of this type generally do not guarantee convergence to a globally optimal solution, and that sets them apart from deterministic methods.

Some parallel variants are proposed for deterministic algorithms of lipschitzian global optimization with guaranteed convergence \cite{RefGergel2005,RefEvtushenko,RefHe,RefPaulavicius}. However, these versions of algorithms are parallelized on CPU; there are no implementations on GPU. For example, parallelization of algorithms based on the ideas of the branch and bound method is carried out for a small number of processors using OpenMP and MPI technologies \cite{RefPaulavicius}. It is noteworthy that the branch and bound method is applied in other areas of optimization; in particular, combinatorial optimization and integer programming. Using GPU shows good results in these areas \cite{RefBoukedjar}, \cite{RefCarneiro}. It is explained by the following factor. Problems of this class (such as the Travelling Salesman Problem or the Knapsack Problem) are characterized by short time of objective function value calculation in one point and, at the same time, relatively long time required to process the results. Therefore, the most labor-intensive operations here are branching and bound computation, which are implemented on GPU (see \cite{RefBoukedjar}, \cite{RefCarneiro}).

This paper considers the problems of lipschitzian global optimization that are frequently encountered in applications. They are characterized by a considerable time required to calculate the function value in one point (in comparison with the processing of the results of such calculation). Therefore, calculation of the value of an optimized function in one point can be done using a GPU, and the main objective of the algorithm (working at CPU) is to efficiently select points of parallel trials. The article contains the results of an investigation into the GPU implementation of the global search parallel algorithm developed according to the information-statistical approach presented in the monograph \cite{RefStrongin2000}.
%new text end

The main part of the paper has the following structure. Section \ref{sec:2} contains a description of the basic algorithm of global search. Section \ref{sec:3} contains a scheme of its parallelization, and section \ref{sec:4} includes theoretical statements that characterize the speedup and non-redundancy of the parallel algorithm. In section \ref{sec:5}, ways of possible implementation of the parallel algorithm on GPU are discussed. Section \ref{sec:6} contains the results of numerical experiments; a comparison of the sequential algorithm and parallel CPU and GPU versions for solving a series of test multidimensional multiextremal problems is also made. Section \ref{sec:7} concludes the paper.

@@

Let us consider the problem of search for a global minimum of an $N$-dimensional function $\varphi(y)$ within a hyper interval $D$
\begin{eqnarray}\label{eq:1}
& \varphi(y^\ast)=\min{\left\{\varphi(y):y\in D\right\}},\\
& D=\left\{y\in R^N: a_i\leq y_i \leq b_i, 1\leq i \leq N\right\}. \nonumber
\end{eqnarray}

@@

By using this kind of mapping it is possible to reduce the multidimensional problem~(\ref{eq:1}) to a one-dimensional problem
\[
\varphi(y^\ast)=\varphi(y(x^\ast))=\min{\left\{\varphi(y(x)): x\in[0,1]\right\}}.
\]
An important property of such mapping is preservation of boundedness of function relative differences  (see \cite{RefStrongin2000}): if the function $\varphi(y)$ in the domain $D$ satisfies the Lipschitz condition, then the function $\varphi(y(x))$ on the interval $[0,1]$ will satisfy a uniform H{\"o}lder condition
\[
\left|\varphi(y(x_1))-\varphi(y(x_2))\right|\leq H\left|x_1-x_2\right|^{1/N},
\]
where the H{\"o}lder constant $H$ is linked to the Lipschitz constant $L$ by the relation
\[
H=4Ld\sqrt{N},\;\; d=\max{\left\{b_i-a_i:1\leq i \leq N\right\}}.
\]

Therefore, it is possible, without limitation of generality, to consider minimization of one-dimensional function
\[
f(x)=\varphi(y(x)), \;\; x\in[0,1],
\]
satisfying the H{\"o}lder condition.
%new text begin
Once again, we recall that the Peano curve $y(x)$ is defined as a limit object emerging in some sequential construction. Appropriate approximations to $y(x)$ are to be used for practical application.
%new text end

@@

At the first iteration of the method the trial is carried out at an arbitrary internal point $x^1$ of the interval $[0,1]$. The point of trial at the next iteration $(k+1)$ is determined according to these rules.

Step 1. Renumber points of the set
\[
X_k=\{x^1,\dots,x^k\}\cup\left\{0\right\}\cup\left\{1\right\},
\]
which includes boundary points of the interval $[0,1]$ and the points of the previous trials, with subscripts in increasing order of coordinate values, i.e.
\[
0=x_0<x_1<\dots <x_k<x_{k+1}=1.
\]

@@

Step 3. Calculate a \textit{characteristic} for every interval $(x_{i-1}, x_i), \; 1\leq i \leq k+1,$   according to the following formulas
\[
R(1)=2\Delta_1-4\frac{z_1}{M},
\]
\begin{equation}\label{eq:14}
R(i)=\Delta_i+\frac{(z_i-z_{i-1})^2}{M^2\Delta_i}-2\frac{z_i+z_{i-1}}{M},1<i<k+1,
\end{equation}
\[
R(k+1)=2\Delta_{k+1}-4\frac{z_k}{M}.
\]

Step 4. Find interval $(x_{t-1},x_t)$ with the maximum characteristic
\[
R(t)=\max{\left\{R(i): 1 \leq i \leq k+1\right\}}.
\]

Step 5. Carry out a trial at the point $x^{k+1}\in(x_{t-1},x_t)$, calculated using the following formulas
\[
x^{k+1} = \frac{x_t+x_{t-1}}{2}, \textrm{ if } t=1 \textrm{ or } t=k+1,
\]
\[
x^{k+1} = \frac{x_t+x_{t-1}}{2} - \mathrm{sign}(z_t-z_{t-1})\frac{1}{2r}\left[\frac{\left|z_t-z_{t-1}\right|}{\mu}\right]^N, \textrm{ if } 1<t<k+1.
\]

@@

%new text begin
Remark 1. The global search algorithm was developed in the framework of a stochastic model (see \cite{RefStrongin2000}). From this point of view the normalized characteristics $R(i)$ from (\ref{eq:14}) can be considered as probabilities of locating the global minimum within the interval  $(x_{i-1},x_i), 1\leq i \leq k+1$. Thus, at each iteration a new trial point is selected inside the interval, which has the greatest probability of finding the global minimum. 
%new text end

Remark 2. The proof of convergence of this algorithm is provided in \cite{RefStrongin2000}. The modifications taking into account existence of inequality constraints in the problem and the information about the objective function derivative are given in \cite{RefBarkalov,RefGergel1996,RefGergel1997}.

Let us illustrate the work of this \textit{global search algorithm} (GSA) using a dimension reduction for minimization of a two-dimensional multiextremal function
\begin{eqnarray} \nonumber \label{eq:19}
\varphi(y)= -&\left\{\left(\sum^{7}_{i=1}\sum^{7}_{j=1}A_{ij}g_{ij}(y)+B_{ij}h_{ij}(y)\right)^2+\right. \\
&\left.\left(\sum^{7}_{i=1}\sum^{7}_{j=1}C_{ij}g_{ij}(y)+D_{ij}h_{ij}(y)\right)^2\right\}^{1/2},\\ \nonumber
\end{eqnarray}
where
\begin{eqnarray} \nonumber
& y=(y_1,y_2)\in R^2, 0 \leq y_s \leq 1, s=1,2, \\ \nonumber
& g_{ij}(y)=\sin(i\pi y_1)\sin(j\pi y_2),  \\ \nonumber
& h_{ij}(y)=\cos(i\pi y_1)\cos(j\pi y_2), \nonumber 
\end{eqnarray}
and coefficients $A_{ij}, B_{ij}, C_{ij}, D_{ij}$  are taken uniformly in the interval $[-1,1]$. The experiment used the method parameters $r=3$, $\epsilon=10^{-3}$, and the evolvent construction parameter $m=12$. Fig.~\ref{fig:1} shows the contour lines of a function generated according to (\ref{eq:19}) and the points of 1103 trials carried out using the method before the required accuracy was obtained.
\begin{figure}
	\center
  \includegraphics[width=0.75\textwidth]{fig2.jpg} 
  \caption{Minimization of a two-dimensional function by the sequential algorithm}
  \label{fig:1}       % Give a unique label
\end{figure}

@@

%new text begin
The third option is to change an algorithm for the purpose of carrying out several trials in parallel. This approach is most promising since it is characterized by efficiency and generality. Firstly, the part of computing process, in which the bulk of calculations is carried out, is parallelized. Secondly, the approach is applicable for a wide class of characteristical algorithms of multiextremal optimization \cite{RefGrishagin1997}.
%new text end

Let us describe a change of rules of global search algorithm for the purpose of simultaneous (synchronous) carrying out in one iteration of method $p$ of trials. Let us make $k(n)$ the total number of trials, carried out after $n$ parallel iterations.

Suppose $n\geq 1$  iterations of the method are carried out, in which trials were performed at $k=k(n)$ points $x^i,1\leq i \leq k$. Then points $x^{k+1},\dots,x^{k+p}$  of search trials of the next $n+1$ iteration are determined according to the rules.

@@

Step 5. Carry out new trials at points $x^{k+j}\in(x_{t_j-1},x_{t_j}), 1\leq j\leq p$, calculated using the formulas
\[
x^{k+j} = \frac{x_{t_j}+x_{t_j-1}}{2}, \textrm{ if } t_j=1 \textrm{ or } t_j=k+1,
\]
\[
x^{k+j} = \frac{x_{t_j}+x_{t_j-1}}{2} - \mathrm{sign}(z_{t_j}-z_{t_j-1})\frac{1}{2r}\left[\frac{\left|z_{t_j}-z_{t_j-1}\right|}{\mu}\right]^N, \textrm{ if } 1<t_j<k+1.
\]

@@

The modifications of the parallel algorithm with the account of the information about the minimized function derivative are given in \cite{RefGergel1999}.

@@

Let us use a parallel algorithm of global search for solving a problem from section \ref{sec:2} with the numbers of parallel trials $p=2$ and $p=4$. With $p=2$ the number of iterations was $569$, and the number of trials was $1138$. With $p=4$ the number of iterations was reduced to $270$, and $1080$ trials were carried out. Fig.~\ref{fig:2} shows contour lines of the function with points of the trials performed by the method with $p=2$. Points of redundant trials carried out using a parallel algorithm are marked in the figure by "+". The concept of redundancy is discussed in more detail in the following section.
\begin{figure}
	\center
  \includegraphics[width=0.75\textwidth]{fig3.jpg} 
  \caption{Minimization of a two-dimensional function by the parallel algorithm}
  \label{fig:2}
\end{figure}

@@

Let us describe theoretical properties of a parallel algorithm, which characterize its speedup. One of the main indicators of efficiency of parallel algorithms (in any field, not only in global optimization) is a speedup in time
\[
S(p)=T(1)/T(p),
\]
where $T(1)$ is the time required to solve a problem by a sequential algorithm, and $T(p)$ is the time for solving the same problem by a parallel algorithm in a system with $p$ computing elements. The characteristic of efficiency of a parallel algorithms (in relation to optimization algorithms) is also a speedup in iterations
\begin{equation}\label{eq:26}
s(p)=n(1)p/n(p),
\end{equation}
where $n(1)$ is the number of trials carried out using the sequential method, and $n(p)$ is the number of trials carried out using the parallel method with $p$ processors. This characteristic is especially important since for solving applied problems the time required to carry out a trial exceeds the time required to process its results.

It is obvious that the number of trials $n(p)$ for sequential and parallel algorithms described in sections \ref{sec:2} and \ref{sec:3} will differ. Actually, the sequential algorithm of global search when selecting point $x^{k+1}$ of the next $k+1$ trial possesses complete information received at the previous $k$ iterations. Based on the same information, the parallel algorithm of global search selects $p$ points $x^{k+j}, 1\leq j \leq p$, rather than one at iteration $k+1$. It means that selection of point $x^{k+j}$ is carried out in absence of information on the trial results at points $x^{k+i}, 1\leq i<j$. Only the first point $x^{k+1}$ will coincide with the point selected by the sequential algorithm. Other trials points generally cannot coincide with the points generated by the sequential algorithm. The efficiency offered by the use of parallel processors can be reduced when such trials are performed. Therefore, let us consider such trials as ``redundant'', and the value 
\[
\lambda(p) = \left\{
   \begin{array}{lr}
     (n(p)-n(1))/n(p), & n(p) > n(1),\\
     0, & n(p)\leq n(1),
   \end{array}
\right.
\]
will represent ``method redundancy''.

@@

\textbf{Theorem}. Suppose $x^\ast$ is the point of global minimum, $x'$ is the point of the local minimum of function $f(x)$, and the following conditions are fulfilled:
\begin{enumerate}
	\item
	Inequality 
	\begin{equation}\label{eq:28}
	f(x')-f(x^\ast)\leq \delta, \delta>0,
	\end{equation}
	holds.
	\item
	The initial $q(l)$ trials of the sequential and parallel methods coincide, i.e.
	\[
	\left\{x^1,\dots,x^{q(l)}\right\}=\left\{y^1,\dots,y^{q(l)}\right\},
	\]
	where 
	\[
	\left\{x^1,\dots,x^{q(l)}\right\}\subset\left\{x^k\right\}, \ \left\{y^1,\dots,y^{q(l)}\right\}\subset\left\{y^m\right\}.
	\]
	\item
	There exists a point $y^n\in \left\{y^m\right\}, n<q(l),$ such that $x'\leq y^n \leq x^\ast$ or \mbox{$x^\ast \leq y^n \leq x'$}.
	\item
	For the value $M$ from (\ref{eq:11}) the following inequality
	\[
	M>2^{2-1/N}H
	\]
	holds, where $H$ is the H{\"o}lder constant of the minimized function.
\end{enumerate}

>>>>

\title{Parallel global optimization on GPU 
	\thanks{The revised version of the paper was supported by the Russian Science Foundation, project No 15-11-30022 ``Global optimization, supercomputing computations, and applications''}
}
%\subtitle{Do you have a subtitle?\\ If so, write it here}
%\titlerunning{Short form of title}        % if too long for running head

@@

\institute{K. Barkalov  \at
              Computational Mathematics and Cybernetics Faculty, Lobachevsky State University of Nizhni Novgorod, Nizhni Novgorod, Russia \\
              Tel.: +7-831-4623356\\
              Fax: +7-831-4623085\\
              \email{barkalov@vmk.unn.ru} 
           \and
           V. Gergel \at
							\email{gergel@unn.ru} 
}

@@

\begin{abstract}
This work considers a parallel algorithm for solving multidimensional multiextremal optimization problems. This algorithm uses Peano-type space filling curves for dimension reduction. Conditions of non-redundant parallelization of the algorithm are considered. Efficiency of the algorithm on modern computing systems with the use of graphics processing units (GPUs) is investigated. Speedup of the algorithm using GPU as compared with the same algorithm implemented on CPU only is demonstrated experimentally. Computational experiments are carried out on a series of several hundred multidimensional multiextremal problems.
\keywords{Global optimization \and Parallel computing \and Graphics processing unit \and Peano-type space filling curves \and Characteristical algorithms}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

This paper considers parallel algorithms for solving multiextremal optimization problems. The objective function in such problems has several local optima (usually their number is not known and can be large). 
%new text begin
In such problems a reliable estimation of the global optimum is possible only with an a priori information about the function, allowing to estimate lower bound of the objective function using known values at search points. 
%new text end
Very often, such information is presented as the assumption that the objective function satisfies a Lipschitz condition with an a priori unknown constant $L$. 
%new text begin
The important feature of the global optimization problems is that the objective function is commonly not given by an analytical formula. The usual way to calculate the function value is to apply a computer-aided algorithm. Thus, each such calculation may be interpreted as running some blackbox device  to produce the outcome for the preset input. These runs could require substantial computer resources and therefore should not be too numerous. 
%new text end
%new word - such
Such problems are common in applications (problems of optimal design of objects and technological processes in various fields of technology, problems of model fitting according to observed data in scientific research, optimization of control processes, etc., see the review \cite{RefPinter}).

Multiextremal optimization problems offer little in terms of analytical solution, while numerical solution is computationally challenging as computational costs often grow exponentially in dimension. Using modern parallel computational systems it is possible to broaden the area of applicability of global optimization methods. At the same time, the problem of efficient parallelization of the global search process arises. Therefore, development of efficient parallel methods for numerical solution of multiextremal optimization problems and implementation of corresponding software for modern computational systems is the obvious focus of research in this area.

One of the promising trends in the field of parallel global optimization (as in many areas connected with software implementation of complex algorithms) is the use of graphics processing units (GPUs). During the last decade, productivity of GPUs has been rapidly increasing to satisfy the continuously growing demand of graphic application developers. Furthermore, some principles of development of graphics hardware have changed over the past few years. Today a GPU is a high-performance flexibly programmable and massively parallel processor that can provide solution to many computationally intensive problems \cite{RefHwu}.

%new text begin
By now, parallel implementations have been suggested for most of the existing sequential algorithms. A review of modern methods of parallel global optimization can be found, for example, in \cite{RefApuzzo}. The methods described in \cite{RefApuzzo} can be divided into two classes: deterministic methods of global optimization, and heuristic methods, which are based one way or another on the random search concept. Algorithms, which belong to the latter class, allow the use of massive parallelism and are easily implemented on GPU (see works \cite{RefFerreiro,RefZhu,RefGarcia,RefMussi} and the review \cite{RefLangdon}). Owing to their stochastic nature, algorithms of this type can guarantee convergence to a globally optimal solution only in probabilistic sense, and that sets them apart from deterministic methods.

Some parallel variants are proposed for deterministic algorithms of Lipschitzian global optimization in \cite{RefGergel2005,RefEvtushenko,RefHe,RefPaulavicius}. However, these versions of algorithms are parallelized on a computing cluster with distributed memory using multi-core CPU; implementations of Lipschitzian global optimization algorithms on GPU have not been suggested yet. For example, parallelization of algorithms based on the ideas of the branch and bound method is carried out for a small number of nodes using OpenMP and MPI technologies \cite{RefPaulavicius}. It is noteworthy that the branch and bound method is applied in other areas of optimization; in particular, combinatorial optimization and integer programming. Using GPU shows good results in these areas \cite{RefBoukedjar,RefCarneiro}. It is explained by the following factor. Problems of this class (such as the Travelling Salesman Problem or the Knapsack Problem) are characterized by short time of objective function value calculation at one point and, at the same time, relatively long time required to process the results. Therefore, the most of labor-intensive operations here consist in branching and bound computations, which are implemented on GPU (see \cite{RefBoukedjar,RefCarneiro}).

This paper considers the problems of Lipschitzian global optimization that are frequently encountered in applications. They are characterized by a considerable time required to calculate the function value at one point (in comparison with the processing of the results of such calculation). For example, the objective function can be defined by solving SLAE, ODE, PDE, or Monte Carlo simulation. Currently, GPU may be used to solve all these types of problems \cite{RefKindratenko}. Therefore, calculation of values of an optimized function can be done using a GPU, and the main objective of the algorithm (working at CPU) is to efficiently select points of parallel trials. The article contains the results of an investigation into the GPU implementation of the global search parallel algorithm developed according to the information-statistical approach presented in the monograph \cite{RefStrongin2000}. Some preliminary results of research in this area have been published in \cite{RefLebedevBarkalov}.
%new text end

The main part of the paper has the following structure. Section \ref{sec:2} contains a description of the basic algorithm of global search. Section \ref{sec:3} describes a scheme of its parallelization, and section \ref{sec:4} includes theoretical statements that characterize the speedup and non-redundancy of the parallel algorithm. In section \ref{sec:5}, ways of possible implementation of the parallel algorithm on GPU are discussed. Section \ref{sec:6} contains the results of numerical experiments; a comparison of the sequential algorithm and parallel CPU and GPU versions for solving a series of test multidimensional multiextremal problems is also made. Section \ref{sec:7} concludes the paper.

@@

Let us consider the problem of search for a global minimum of an $N$-dimensional function $\varphi(y)$ within a hyperinterval $D$
\begin{eqnarray}\label{eq:1}
& \varphi(y^\ast)=\min{\left\{\varphi(y):y\in D\right\}},\\
& D=\left\{y\in R^N: a_i\leq y_i \leq b_i, 1\leq i \leq N\right\}. \nonumber
\end{eqnarray}

@@

By using this kind of mapping it is possible to reduce the multidimensional problem~(\ref{eq:1}) to a univariate problem
\[
\varphi(y^\ast)=\varphi(y(x^\ast))=\min{\left\{\varphi(y(x)): x\in[0,1]\right\}}.
\]
An important property of such mapping is preservation of boundedness of function relative differences  (see \cite{RefStrongin2000}): if the function $\varphi(y)$ in the domain $D$ satisfies the Lipschitz condition, then the function $\varphi(y(x))$ on the interval $[0,1]$ will satisfy a uniform H{\"o}lder condition
\[
\left|\varphi(y(x_1))-\varphi(y(x_2))\right|\leq H\left|x_1-x_2\right|^{1/N},
\]
where the H{\"o}lder constant $H$ is linked to the Lipschitz constant $L$ by the relation
\[
H=4Ld\sqrt{N},\;\; d=\max{\left\{b_i-a_i:1\leq i \leq N\right\}}.
\]

Therefore, it is possible, without loss of generality, to consider minimization of univariate function
\[
f(x)=\varphi(y(x)), \;\; x\in[0,1],
\]
satisfying the H{\"o}lder condition.
%new text begin
Once again, we recall that the Peano curve $y(x)$ is defined as a limit object emerging in some sequential construction. Appropriate approximations to $y(x)$ are to be used for practical application.
%new text end

@@

At the first iteration of the method the trial is carried out at an arbitrary internal point $x^1$ of the interval $[0,1]$. The point of trial at the next iteration $(k+1)$ is determined according to the rules presented below.

Step 1. Renumber points of the set
\[
X_k=\{x^1,\dots,x^k\}\cup\left\{0\right\}\cup\left\{1\right\},
\]
which includes boundary points of the interval $[0,1]$ and the points of the previous trials, with subscripts in increasing order of coordinate values, i.e.,
\[
0=x_0<x_1<\dots <x_k<x_{k+1}=1.
\]

@@

Step 3. Calculate a \textit{characteristic} for every interval $(x_{i-1}, x_i), \; 1\leq i \leq k+1,$   according to the following formulae
\[
R(1)=2\Delta_1-4\frac{z_1}{M},
\]
\begin{equation}\label{eq:14}
R(i)=\Delta_i+\frac{(z_i-z_{i-1})^2}{M^2\Delta_i}-2\frac{z_i+z_{i-1}}{M},1<i<k+1,
\end{equation}
\[
R(k+1)=2\Delta_{k+1}-4\frac{z_k}{M}.
\]

Step 4. Find interval $(x_{t-1},x_t)$ with the maximum characteristic
\begin{equation}\label{eq:141}
R(t)=\max{\left\{R(i): 1 \leq i \leq k+1\right\}}.
\end{equation}

Step 5. Carry out a trial at the point $x^{k+1}\in(x_{t-1},x_t)$, calculated using the following formulae
\[
x^{k+1} = \frac{x_t+x_{t-1}}{2}, \textrm{ if } t=1 \textrm{ or } t=k+1,
\]
\begin{equation}\label{eq:142}
x^{k+1} = \frac{x_t+x_{t-1}}{2} - \mathrm{sign}(z_t-z_{t-1})\frac{1}{2r}\left[\frac{\left|z_t-z_{t-1}\right|}{\mu}\right]^N, \textrm{ if } 1<t<k+1.
\end{equation}

@@

%new text begin
Remark 1. The global search algorithm was developed in the framework of a stochastic model (see \cite{RefStrongin2000}). From this point of view the normalized characteristics $R(i)$ from (\ref{eq:14}) can be considered as probabilities of locating the global minimum within the interval  $(x_{i-1},x_i), 1\leq i \leq k+1$. Thus, at each iteration a new trial point is selected inside the interval, which has the greatest probability of finding the global minimum.

Remark 2. In a different way, the global search algorithm has a geometric interpretation (see \cite{RefSergeyev2001}). Using trial points it is possible to design a support function (\textit{minorant}) for which characteristic $R(i)$ is the minimum value of the minorant in the interval $(x_{i-1},x_i), 1\leq i\leq k+1$, taken with the opposite sign. Then according to (\ref{eq:141}) new trial is carried out in the interval where the global minimizer of the current minorant is situated and the new trial point $x_{k+1}$ from (\ref{eq:142}) coincides with it.

%new text end

Remark 3. The proof of convergence of this algorithm is provided in \cite{RefStrongin2000}. The modifications taking into account existence of inequality constraints in the problem and the information about the objective function derivative are given in \cite{RefBarkalov,RefGergel1996,RefGergel1997}.

Let us illustrate the work of this \textit{global search algorithm} (GSA) using a dimension reduction for minimization of a multiextremal function of two variables 
\begin{eqnarray} \nonumber \label{eq:19}
\varphi(y)= -&\left\{\left(\sum^{7}_{i=1}\sum^{7}_{j=1}A_{ij}g_{ij}(y)+B_{ij}h_{ij}(y)\right)^2+\right. \\
&\left.\left(\sum^{7}_{i=1}\sum^{7}_{j=1}C_{ij}g_{ij}(y)+D_{ij}h_{ij}(y)\right)^2\right\}^{1/2},\\ \nonumber
\end{eqnarray}
where
\begin{eqnarray} \nonumber
& y=(y_1,y_2)\in R^2, 0 \leq y_s \leq 1, s=1,2, \\ \nonumber
& g_{ij}(y)=\sin(i\pi y_1)\sin(j\pi y_2),  \\ \nonumber
& h_{ij}(y)=\cos(i\pi y_1)\cos(j\pi y_2), \nonumber 
\end{eqnarray}
and coefficients $A_{ij}, B_{ij}, C_{ij}, D_{ij}$  are taken uniformly in the interval $[-1,1]$. The experiment used the method parameters $r=3$, $\epsilon=10^{-3}$, and the evolvent construction parameter $m=12$. Fig.~\ref{fig:1} shows the level lines of a function generated according to (\ref{eq:19}) and the points of 1103 trials carried out using the method before the required accuracy was obtained.
\begin{figure}
	\center
  \includegraphics[width=0.75\textwidth]{fig2.jpg} 
  \caption{Minimization of a function of two variables by the sequential algorithm}
  \label{fig:1}       % Give a unique label
\end{figure}

@@

%new text begin
The third option is to change an algorithm for the purpose of carrying out several trials in parallel. This approach is most promising since it is characterized by efficiency and generality. Firstly, the part of computing process, in which the bulk of calculations is carried out, is parallelized. Secondly, the approach is applicable for a wide class of characteristical algorithms of multiextremal optimization \cite{RefSergeyev1994,RefSergeyevGrishagin,RefGrishagin1997,RefGrishagin2001}.
%new text end

Let us describe a change of rules of global search algorithm for the case when the algorithm carries out during one iteration $p>1$ trials simultaneously (each trial on a separate processor). Let $k(n)$ be the total number of trials, carried out after $n$ parallel iterations.

Suppose $n\geq 1$  iterations of the method have been carried out, in which trials were performed at $k=k(n)$ points $x^i,1\leq i \leq k$. Then points $x^{k+1},\dots,x^{k+p}$  of search trials of the next $n+1$ iteration are determined according to the following rules.

@@

Step 5. Carry out new trials at points $x^{k+j}\in(x_{t_j-1},x_{t_j}), 1\leq j\leq p$, calculated using the formulae
\[
x^{k+j} = \frac{x_{t_j}+x_{t_j-1}}{2}, \textrm{ if } t_j=1 \textrm{ or } t_j=k+1,
\]
\[
x^{k+j} = \frac{x_{t_j}+x_{t_j-1}}{2} - \mathrm{sign}(z_{t_j}-z_{t_j-1})\frac{1}{2r}\left[\frac{\left|z_{t_j}-z_{t_j-1}\right|}{\mu}\right]^N, \textrm{ if } 1<t_j<k+1.
\]

@@

The modifications of the parallel algorithm which takes into account the information about the minimized function derivative are given in \cite{RefGergel1999}.

@@

Let us use a parallel algorithm of global search for solving a problem from section \ref{sec:2} with the numbers of parallel trials $p=2$ and $p=4$. With $p=2$ the number of iterations was $569$, and the number of trials was $1138$. With $p=4$ the number of iterations was reduced to $270$, and $1080$ trials were carried out. Fig.~\ref{fig:2} shows level lines of the function with points of the trials performed by the method with $p=2$. Points of redundant trials carried out using a parallel algorithm are marked in the figure by "+". The concept of redundancy is discussed in more detail in the following section.
\begin{figure}
	\center
  \includegraphics[width=0.75\textwidth]{fig3.jpg} 
  \caption{Minimization of a function of two variables by the parallel algorithm}
  \label{fig:2}
\end{figure}

@@

Let us describe theoretical properties of a parallel algorithm, which characterize its speedup. One of the main indicators of efficiency of parallel algorithms (in any field, not only in global optimization) is a speedup in time
\[
S(p)=T(1)/T(p),
\]
where $T(1)$ is the time required to solve a problem by a sequential algorithm, and $T(p)$ is the time for solving the same problem by a parallel algorithm in a system with $p$ computing elements. The characteristic of efficiency of parallel algorithms (in relation to optimization algorithms) is also a speedup in iterations
\begin{equation}\label{eq:26}
s(p)=n(1)p/n(p),
\end{equation}
where $n(1)$ is the number of trials carried out by the sequential method, and $n(p)$ is the number of trials carried out by the parallel method with $p$ processors. This characteristic is especially important since for solving applied problems the time required to carry out a trial exceeds the time required to process its results.

It is obvious that the number of trials $n(p)$ for sequential and parallel algorithms described in sections \ref{sec:2} and \ref{sec:3} will differ. Actually, the sequential algorithm of global search when selecting point $x^{k+1}$ of the next $k+1$ trial possesses complete information received at the previous $k$ iterations. Based on the same information, the parallel algorithm of global search selects $p$ points $x^{k+j}, 1\leq j \leq p$, rather than one at iteration $k+1$. It means that selection of point $x^{k+j}$ is carried out in absence of information on the trial results at points $x^{k+i}, 1\leq i<j$. Only the first point $x^{k+1}$ will coincide with the point selected by the sequential algorithm. Other trials points generally cannot coincide with the points generated by the sequential algorithm. The efficiency owing to the use of parallel processors can be reduced when such trials are performed. Therefore, let us consider such trials as ``redundant'', and the value 
\[
\lambda(p) = \left\{
   \begin{array}{lr}
     (n(p)-n(1))/n(p), & n(p) > n(1),\\
     0, & n(p)\leq n(1),
   \end{array}
\right.
\]
will represent ``method redundancy''.

@@

\textbf{Theorem}. Suppose $x^\ast$ is the point of global minimum, $x'$ is the point of the local minimum of function $f(x)$, and the following conditions are fulfilled:
\begin{enumerate}
	\item
	Inequality 
	\begin{equation}\label{eq:28}
	f(x')-f(x^\ast)\leq \delta, \delta>0,
	\end{equation}
	holds.
	\item
	The initial $q(l)$ trials of the sequential and parallel methods coincide, i.e.,
	\[
	\left\{x^1,\dots,x^{q(l)}\right\}=\left\{y^1,\dots,y^{q(l)}\right\},
	\]
	where 
	\[
	\left\{x^1,\dots,x^{q(l)}\right\}\subset\left\{x^k\right\}, \ \left\{y^1,\dots,y^{q(l)}\right\}\subset\left\{y^m\right\}.
	\]
	\item
	There exists a point $y^n\in \left\{y^m\right\}, n<q(l),$ such that $x'\leq y^n \leq x^\ast$ or \mbox{$x^\ast \leq y^n \leq x'$}.
	\item
	For the value $M$ from (\ref{eq:11}) the following inequality
	\[
	M>2^{2-1/N}H
	\]
	holds, where $H$ is the H{\"o}lder constant of the minimized function.
\end{enumerate}

Then a parallel algorithm of global search using two processors will be non-redundant (i.e., $s(2)=2$, $\lambda(2)=0$) while the following condition is satisfied 
\begin{equation}\label{eq:32}
\left(x_{t_j}-x_{t_j-1}\right)^{1/N} > \frac{4\delta}{M-2^{2-1/N}H}, 1\leq j\leq p,	
\end{equation}
where $t_j$ are determined according to (\ref{eq:21}).
